Testing and Isolation Efficacy: Insights from a Simple Epidemic Model

### 2021-02-09 SIR_manuscript.tex
abstract revised, some David's comments were addressed

The potential advantage of a slow rate of test return on the dynamics of an epidemic is real, often neglected, and needs to be quantified. At the individual level, this advantage can manifest in the following sense: individuals awaiting their test results or who have tested positive may partially or fully self-isolate, thus reducing or eliminating their potential in the transmission process.

At the community-level (e.g., nursing homes, LTC facilities, etc.), fast test reporting may be more beneficial. It gives a rapid assessment of the situation, identifies hot spots, and may enable rapid contact-tracing.  
In this paper, we investigated this individual-level effect of testing processes on the epidemic dynamics by developing a SIR-type model.

Although the model development was motivated by the \covid epidemic, the model has general epidemiological and testing structures. The realistic components of the model include testing intensity, test sensitivity and specificity, rate of test return, and isolation. The novel component is the compartment-specific relative testing weights, which reflect the testing strategies-- surveillance, diagnosis, or control.
Here, we showed that the slow test reporting can be beneficial on the dynamics of an epidemic. In particular, it is possible for the basic reproduction number, $\Rnum$, to be increasing with respect to the rate of test return.

Further, we compare two testing strategies, random vs. targeted, in this framework. We conclude that the targeted testing strategy is more effective than the random one in the sense that a lower range of testing intensity is required to keep $\Rnum <1$. Also, targeted testing reduces the individual-level of advantage of a slow rate of test return.

\fady{General thoughts on the abstract (take with a grain of salt): I think the abstract is eye-catching and interesting. I think leading in by highlighting potential advantages to a slow test return may be ``provocative" (which may not necessarily be bad, I'm not sure). I think the second paragraph in some sense undermines or clashes with the first paragraph, especially since we don't actually address these potential advantages in this model. Maybe one way to spin this around would be to lead in with a discussion of the advantages of a FAST test return (i.e., the second paragraph), and then say ``but situations where returning tests more rapidly is BAD have not been studied". I think that could have better harmony or coordination.}

@@

Because individuals who have been tested or received a positive test are likely to self-isolate---reducing or eliminating  their potentially infectious contacts---the dynamics of an epidemic will depend on who gets tested and
why. For a given testing intensity (tests performed per day), testing will have a stronger effect on infection rates
the more strongly it is focused on people who are infectious.\david{deleted `` for SARS-CoV-2''.  Since this is an idealized model intended for insights, it is probably better to use more generic language most of the time and not say covid unless necessary in a specific context.}
This level of focus depends in turn on the purpose and design of testing programs.
(1) When testing is done for the purposes of disease surveillance \citep{foddai2020base}
we might expect tests to be assigned randomly across the population (in an ideal world: some bias in selection for testing is probably inevitable), although a stratified sampling design would be preferred for statistical efficiency \citep{graubard1996modelling} \ali{better ref?}.

Over the course of the \covid pandemic, however, the vast majority of testing has been done with other goals --
primarily diagnosis (determining COVID19 infection status for clinical purposes), or control (determining COVID19 status in order to quickly isolate cases that have been found by contact tracing), which we characterize as \emph{targeted} testing strategies.
In these cases, testing probabilities vary widely across epidemiological compartments; in our dynamical model, we will characterize these probabilities by assigning a \percap testing weight to each compartment that determines the \emph{relative} probability that an individual in that compartment will be selected for testing (see Methods).

When testing is used primarily for diagnosis it will focus on people with COVID19-like symptoms; thus the testing weights for people with \david{escape spaces after dots unless it is a full stop}vs.\ without SARS-CoV-2 infections will depend on the relative magnitudes of the probability of having COVID19-like symptoms when infected with SARS-CoV-2 (i.e., the relative proportions of presymptomatic or asymptomatic vs.\ symptomatic infections) and the probability of having COVID19-like symptoms when \emph{not} infected (e.g., due to other respiratory illnesses). Testing for epidemic control will focus on people who are known to have been in contact with known COVID19 cases; in this case the testing weights for infected vs.\ uninfected people will depend on the probability of infection given contact, as well as the thoroughness and effectiveness of the system for identifying suspicious contacts.

In addition to the overall testing intensity, the testing weights for different compartments, and the degree of self-isolation of test-pending and test-confirmed people, our models will also need to include test specificity and sensitivity\david{I presume you mean that models would ideally take account of these things, which should not be mentioned unless they are defined since the terms are often misunderstood}, as well as the distribution of times from testing to reporting. 
Here, we focus on the the effect of testing intensity, different levels of testing ``focus'' (from random to highly targeted), and rate of test processing on a single index of epidemiological dynamics, the basic reproduction number $\Rnum$.

The early dynamics of an epidemic are determined by $\Rnum$. This index is defined as the expected number of secondary infections arising from a typical infective individual in a completely susceptible population \citep{dietz1993estimation}. 
In the early stages of an epidemic the number of infected individuals is expected grow exponentially over time when $\Rnum>1$. 
To control the spread of disease in the population, \david{the effective} $\Rnum$ must be $<1$.
Although the value of $\Rnum$ cannot completely characterize the dynamics of even the simplest epidemic model
\citep{shaw2021what}, it does give a simple and widely accepted index for the difficulty of control, as well as some indication of the likely final size of an epidemic \citep{ma2006generality}.
\david{This paragraph does not flow from the rest of the intro, and is a weird way to end when the point of the paper is about testing.}

\david{There is some back and forth between generic discussion and COVID-specific comments.  I think COVID should be used as motivation, but that the whole paper should be framed generically.}

@@

We developed a deterministic model, Eqs. \eqref{model}\david{relegating the equations to supp is not OK.   You don't want to publish a paper in which the model is specified exactly only in online supplementary material.  I don't think there is any need for supp material.  You can have appendices, but I would keep everything as part of the paper.}, which groups individuals based on disease status and testing status. Disease states include Susceptible, Infectious and Recovered (thus this is an SIR-type model), and testing status categorizes people as \emph{untested}, waiting-for-\emph{positive}, waiting-for-\emph{negative}, or \emph{confirmed positive} (Figure~\ref{fig:flowchart}). Two `accumulator' compartments, $N$ and $P$, were also incorporated in the model in order to collect cumulative reported negative or positive tests. The model and details of calculation of the basic reproduction number $\Rnum$ are presented in the Supplementary Materials.

@@

In order to link to more applied models\david{what does that mean?  what is an ``applied model''?}, we constructed this model so we could specify the total \percap testing rate. We do this by defining the weighted size of the testing pool $W = W_S S_u + W_I I_u + W_R R_u$, and calculating a scaling parameter for testing as:
\begin{equation}
\label{sigma}
\sigma = \frac{\rho N_0}{W}.
\end{equation}
Thus, the \percap testing rate for compartment $Z$ is $F_Z=\sigma W_Z$, where $Z \in \{S,I,R\}$.  \david{For a generic compartment, I would suggest $X$ rather than $Z$, since $Z$ is often the final size.}  For a high-sensitivity test, infected people typically flow through to the ``confirmed positive'' ($I_c$, $R_c$) compartments and are thus unavailable for further testing.  Over the course of the epidemic, a fixed testing rate as specified in \eqref{sigma} can (if large enough) exhaust the pool of people available for testing, leading to a singularity when no one is left untested.  Although this phenomenon does not affect our analysis of $\Rnum$, it can affect the temporal dynamics (we discuss an adjustment to the model that solves this problem in the Supplementary material).

\david{A comment on notation:  You haven't actually defined $S_u$, $I_u$ etc.  In any case, I urge you to use macros and use Roman font for subscripts that are textual as opposed to variables.  So $S_{\rm u}$ and $I_{\rm p}$ etc.  Then it is clear the subscript isn't a variable with a value.}

The classical SIR model is based on the following implicit assumptions; well-mixed population, homogeneity of the population\david{What's the difference btw ``well-mixed'' and ``homoegeneous''?}, exponentially distributed duration of infection and large population size (see, e.g., \cite{keeling2011modeling}). In addition to these standard assumptions, our model, \ref{model}, assumes: (i) there is a single force of infection, $\Lambda$, defined as follows
\begin{equation}
\label{Lambda}
\Lambda=\beta \frac{(I_u+\eta_w I_n+\eta_w I_p+ \eta_c I_c)}{N_0},
\end{equation}
across all susceptible pools (see Table~\ref{tab:params} for detail\david{there are no details in the table; you need to do a better job of explaining what's going on}), (ii) $\eta_c \leq \eta_w$, i.e., the individuals awaiting test results have a higher transmission probability than the reported individuals. For this analysis, we also assume a perfectly specific test ($p_S=0$). This last assumption combined with the assumption that no individual is in waiting-for-\emph{positive} or \emph{confirmed positive} compartments, i.e., $S_p(0)=S_c(0)=0$, reduces the model to 10 equations (equations c and d  in \eqref{model} are eliminated).

@@

The Disease-Free Equilibrium (DFE) for the SIR model, Eqs. \eqref{model}, is given by solving the coupled system including $S_u+S_n=N_0$ and $F_S S_u-\omega S_n=0$.
\david{you mean by solving the algebraic equations obtain by setting the DEs to 0.  not clear why you bother to mention two of these algebraic eqns.}
The DFE is
\david{don't put blank lines around equations unless you are started a new paragraph.  comment out blank lines if want the .tex file to be more readable}
%%
\begin{equation}
\label{dfe}
S_n^*= \frac{\rho}{\omega} N_0, \ S_u^*= N_0-S_n^*, \text{~and~} I_j=R_j=0 \ \text{for all }j.
\end{equation}
\david{above you had j in Roman font even though it *is* a variable, so I changed to $j$}

The basic reproduction number, $\Rnum$, was calculated by using the next generation matrix method developed by \cite{van2002reproduction}. $\Rnum$ is

\begin{equation}
\label{R0}
\Rnum= (A \times S_u^* + B \times S_n^*) \times C, 
\end{equation}
where
\begin{equation}
\begin{aligned}
\label{eq:abc}
A=& \gamma(\omega+\gamma) + (\gamma \eta_w + \omega \eta_c p_I) F_I, \\
B=& \big(\omega+(F_I+\gamma)\eta_w\big) \gamma+\frac{(\eta_w \gamma+ \eta_c\omega) \omega p_I F_I }{\omega+\gamma}, \\ 
C=& \frac{\beta/\gamma}{N_0 (\gamma(\omega+\gamma)+F_I(\gamma+\omega p_I))}.
\end{aligned}
\end{equation}
\david{put $\beta/\gamma$ in front of the expression in eq \eqref{R0}, so it is clear that the rest is a modification of the the simplest expression.  It would be worth spending some time trying to write the expression in a way that helps the reader understand where the different factors come from.}
Further details of derivation of $\Rnum$ are provided in the Supplementary material section.

@@

The explicit formula for the basic reproduction number, $\Rnum$ \eqref{R0}, provides an opportunity to study the influence of changes in the underlying parameters on the critical index of epidemic dynamics. We are interested in understanding the effect of parameters that can be realistically controlled by changing in isolation, \percap testing intensity and test resulting, i.e., $\eta_c$ and $\eta_w$, $\rho$ and $\omega$, respectively. The following Proposition is the direct result of taking partial derivative of $\Rnum$ \eqref{R0} with respect to selected parameters. Note that we used Taylor approximation of $\Rnum$ at $\rho=0$ due to the complexity of the expression and thus inconclusive.\david{in that case it would be good to establish numerically that the relationships given in the proposition are valid more generally, at least for reasonable parameter ranges}
See the Supplementary material for details.

\david{make sure to use $\Rnum$ not $R_0$ below (and often in the supp material section).  More importantly, you really need to state in words what this proposition is saying.  The reader has to check the param table and think it through themselves.}
\begin{proposition}
\label{prop1}
Using the expression of $\Rnum$ \eqref{R0},
\begin{enumerate}
\item \label{p1:eta}
$\partial{R_0}/\partial{\eta_c} \geq 0$ and $\partial{R_0}/\partial{\eta_w} \geq 0$. 
\item \label{p1:rho}
$\partial{\Rnum}/\partial{\rho} \leq 0$ when $\rho \approx 0$.
\item \label{p1:omega}
$\partial{R_0}/\partial{\omega}$ can be positive or negative when $\rho \approx 0$.
\end{enumerate}
\end{proposition}

@@

We incorporated the compartment-specific relative testing weights, $W_S$, $W_I$ and $W_R$, to model random testing and \emph{targeted} testing strategies. Here, in the case of targeted testing and for the simplicity and illustration purposes, we assumed that infected and recovered individuals are tested at three times the \percap rate of susceptible individuals, thus $W_I/W_S=3$ and $W_R=W_I$. Note that we have not specified a methodology to assign particular relative testing weights corresponding to a particular targeted testing scenario. 
This part needs to be developed further in future work. 
Modeling different targeted testing strategies, equivalently test-specific testing weights in our framework, requires prior information of the conditional probabilities of getting tested for people in a given compartment. 
This can be implied when we would like to quantify and compare the effect of different levels of test forcus \david{??} for infectious people on the basic reproduction number $\Rnum$, and conclude about the disease spread management. For example, when people are tested for "screening", the individuals with potential higher mobility, eg. people who are getting on flights, get more tested and thus the coresponding heavier testing weight is assigned than people awaiting for a surgery and are probably going to stay in a long-term care facility and consequently less mobile and more isolated to begin with. With our model, we would be able to compare the sensitivity of the disease dynamics, through $\Rnum$, with respect to testing processes.

{\bf The \percap testing intensity, $\rho$;}
Proposition \ref{prop1} part \ref{p1:rho} and also Figure \ref{pan} indicates that increasing testing intensity $\rho$ reduces $\Rnum$. This is sensible since as people are moved to test compartments, namely $Z_p$ and $Z_n$ for $Z \in \{S,I,R\}$, they may partially or fully self-isolate. Also individuals who have tested positive, namely $Z_c$ compartment for $Z \in \{S,I,R\}$, are highly likely to self-isolate. The higher probability of being subject to isolation, the lower the force of infection \eqref{Lambda} and the lower $\Rnum$ will become. While our simulation supported this result, analytically it appeared to be hard to conclude this due to the complexity of $\Rnum$ expression.

{\bf The potential advantage of slow test reporting, or favorable-delay-reporting;}
Individuals are highly likely to fully self-isolate when they are either awaiting the test results or they are reported, thus reducing or eliminating their potentially infectious contacts. Thus, the faster the test reporting rate, $\omega$, the shorter these individuals stay in the ``\emph {safe}'' awaiting-confirmed compartments, namely $Z_n$, $Z_p$ and $Z_c$ for $Z \in \{S,I,R\}$, and the more they get involved in the infection process.
This advantage of slow test reporting is real, and neglected. 
We also compare to an individual-level advantage of fast tests: people who test positive may be even more careful.
In our model analysis, Proposition \ref{prop1} part \ref{p1:omega} is describing this potential advantage. 
It states that returning test results more rapidly, i.e., increasing the rate $\omega$, does not necessarily lower the reproduction number $\Rnum$; whether increasing $\omega$ lowers $\Rnum$ depends on the precise combination of model parameters  including test reporting rate, testing strategies represented by compartment-specific testing weights, test sensitivity and specificity, and the level of isolation. 
Specifically, in the case of perfect isolation, i.e., when $\eta_w=0$ and $\eta_c=0$, $\Rnum$ may increase as the test reporting process becomes faster. This can be seen from expression \eqref{Rom}.
Another example of this favorable delay in $\omega$ could be when the test being employed produces many false negatives. Because many infected asymptomatic individuals will believe they are uninfected/uninfectious, thus may unknowingly spread the virus to many others. Again the delay in the test reporting rate keeps these individuals in the ``\emph {safe}'' awaiting-confirmed compartments.

we are missing out on
community-level advantages of fast testing: better assessment of the
situation, identification of hot spots, contact-tracing, etc.
The \percap testing intensity of CIVID-19 after about a year from the first case reported in December 2019, is still low (($\rho \approx 0$) in our model). In near future new test kits may be widely accessible, our model provide insights in this case. 
In particular, if a cheap test can identify on average more infected individuals as an expensive test, then our model predicts that the cheap test will lower $\Rnum$ more. In contrast, if an expensive test can identify on average more infected individuals, it will not necessarily lower $\Rnum$ more than the cheap test. The use of tests cheaper than RT-PCR has been proposed as a potential strategy for containing the COVID-19 pandemic. While cheaper tests may be less sensitive and reliable than RT-PCR, they allow for broader and more intense testing. Using our Taylor approximation of $\Rnum$ near $\rho = 0$, we examined what circumstances (i.e., model parameters) make the use of one test more favourable than another, and give a complete description through inequality \ref{eq:rho1vsrho2}. In general, we found that the expensive test tends to more effectively lower $\Rnum$ when (a) individuals who test positive self-isolate much more than individuals who are waiting for their test result, (b) the time it takes to return tests is much shorter than the mean infectious period, and (c) the testing intensity is much greater for infected individuals than susceptible individuals.

\ali{ not sure if we want these 2 following parageraphs, or shrink it into afew short sentences?}
In addition to the favorable-delay-reporting observation of our model \eqref{model}, the model enables us to quantify the amount of delay required in the test reporting process as a strategy to reduce $\Rnum$.
To give a biological interpretation, we describe the qualitative trends predicted by the inequality \eqref{eq:necsuf} \todo{Fady please confirm}\fady{yes, looks good to me}. To summarise, returning test results more rapidly tends to be favourable (i.e., reduces $\Rnum$) when (i) \emph{Confirmed-positive} individuals, lower their contact much more than individuals who are waiting for their test results (i.e., $\eta_c \ll \eta_w$), (ii) the test is highly sensitive (i.e., $p_I$ is close to 1), and (iii) the targeted testing strategy is used (i.e., $W_i \gg W_s$).

@@

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage
% \widetext
\begin{center}
\textbf{\large Supplemental Materials: Title for main text}
\end{center}
%%%%%%%%%% Merge with supplemental materials %%%%%%%%%%
%%%%%%%%%% Prefix a "S" to all equations, figures, tables and reset the counter %%%%%%%%%%
\setcounter{equation}{0}
\setcounter{figure}{0}
\setcounter{table}{0}
\setcounter{page}{1}
\makeatletter
\renewcommand{\theequation}{S\arabic{equation}}
\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\bibnumfmt}[1]{[S#1]}
\renewcommand{\citenumfont}[1]{S#1}
%%%%%%%%%% Prefix a "S" to all equations, figures, tables and reset the counter %%%%%%%%%%

@@

The use of tests cheaper than RT-PCR has been proposed as a potential strategy for containing the COVID-19 pandemic. While cheaper tests may be less sensitive and reliable than RT-PCR, they allow for broader and more intense testing. In the analysis below, we compare the $\Rnum$ predicted by our model depending on the testing strategy.

>>>>

The effect of testing processes, including (testing and test reporting) on an epidemic dynamics (infection and recovery) can be studied at the individual level or the community level (e.g., nursing homes, LTC facilities, etc.). 
Gaining insights to determine the sensitivity of the epidemic dynamics with respect to the testing processes will depend on underlying factors including the level of focus (individual or community), assumptions (model), and the interplay between these factors. 
In particular, the fast test reporting may be beneficial at the community-level, supported by many studies, as it gives a rapid assessment of the situation, identifies hot spots, and may enable rapid contact-tracing. However, the potential advantage of a slow rate of test return on the dynamics of an epidemic is real, often neglected, and needs to be quantified. At the individual level, this advantage can manifest in the following sense: individuals awaiting their test results or who have tested positive may partially or fully self-isolate, thus reducing or eliminating their potential in the transmission process.
In this paper, we investigated this individual-level effect of testing processes on the epidemic dynamics by developing a SIR-type model.
Although the model development was motivated by the \covid epidemic, the model has general epidemiological and testing structures. The realistic components of the model include testing intensity, test sensitivity and specificity, rate of test return, and isolation. The novel component is the compartment-specific relative testing weights, which reflect the testing strategies-- surveillance, diagnosis, or control.
Here, we showed that the slow test reporting can be beneficial on the dynamics of an epidemic. In particular, it is possible for the basic reproduction number, $\Rnum$, to be increasing with respect to the rate of test return.  
Further, we compare two testing strategies, random vs. targeted, in this framework. We conclude that the targeted testing strategy is more effective than the random one in the sense that a lower range of testing intensity is required to keep $\Rnum <1$. Also, targeted testing reduces the individual-level of advantage of a slow rate of test return.

@@

Because individuals who have been tested or received a positive test are likely to self-isolate---reducing or eliminating  their potentially infectious contacts---the dynamics of an epidemic will depend on who gets tested and
why. For a given testing intensity (tests performed per day), testing will have a stronger effect on infection rates
the more strongly it is focused on people who are infectious.
This level of focus depends in turn on the purpose and design of testing programs.
(1) When testing is done for the purposes of disease surveillance \citep{foddai2020base}
we might expect tests to be assigned randomly across the population (in an ideal world: some bias in selection for testing is probably inevitable), although a stratified sampling design would be preferred for statistical efficiency \citep{graubard1996modelling} \ali{a better ref?}.

Over the course of the \covid pandemic, however, the vast majority of testing has been done with other goals --
primarily diagnosis (determining the infection status for clinical purposes), or control (determining the infection status in order to quickly isolate cases that have been found by contact tracing), which we characterize as \emph{targeted} testing strategies.
In these cases, testing probabilities vary widely across epidemiological compartments; in our dynamical model, we will characterize these probabilities by assigning a \percap testing weight to each compartment that determines the \emph{relative} probability that an individual in that compartment will be selected for testing (see Methods).

When testing is used primarily for diagnosis it will focus on people with infection-like symptoms; thus the testing weights for people with vs.\ without the infections will depend on the relative magnitudes of the probability of having the infection-like symptoms when infected. Specifically, for \covid infection, the testing weights will depend on the relative proportions of presymptomatic or asymptomatic vs.\ symptomatic infections and the probability of having \covid-like symptoms when \emph{not} infected (e.g., due to other respiratory illnesses). Testing for epidemic control will focus on people who are known to have been in contact with known infected cases; in this case the testing weights for infected vs.\ uninfected people will depend on the probability of infection given contact, as well as the thoroughness and effectiveness of the system for identifying suspicious contacts.

In order to understand the effect of testing processes on an epidemic dynamics, we deveopled a mechanistic model with epidemic and testing components. Here, we focus on the the effect of testing intensity, different levels of testing ``focus'' (from random to highly targeted), and rate of test return on a single index of epidemiological dynamics, the basic reproduction number $\Rnum$, which is useful to estimate the early dynamics of an epidemic. This index is defined as the expected number of secondary infections arising from a typical infective individual in a completely susceptible population \citep{dietz1993estimation}. 
In the early stages of an epidemic the number of infected individuals is expected grow exponentially over time when $\Rnum>1$. 
To control the spread of disease in the population, \david{the effective} $\Rnum$ must be $<1$.
Although the value of $\Rnum$ cannot completely characterize the dynamics of even the simplest epidemic model
\citep{shaw2021what}, it does give a simple and widely accepted index for the difficulty of control, as well as some indication of the likely final size of an epidemic \citep{ma2006generality}.
Our model provides insights to the sensitivity of the epidemiological dynamics, through $\Rnum$, with respect to the undelying testing processes.
\david{This paragraph does not flow from the rest of the intro, and is a weird way to end when the point of the paper is about testing.}\ali{edited, but still needs work}

\david{There is some back and forth between generic discussion and \covid-specific comments.  I think COVID should be used as motivation, but that the whole paper should be framed generically.}

@@

We developed a deterministic model, Eqs. \eqref{model}, which groups individuals based on disease status and testing status. Disease states include Susceptible, Infectious and Recovered (thus this is an SIR-type model), and testing status categorizes people as \emph{untested}, waiting-for-\emph{positive}, waiting-for-\emph{negative}, or \emph{confirmed positive} (Figure~\ref{fig:flowchart}). Two `accumulator' compartments, $N$ and $P$, were also incorporated in the model in order to collect cumulative reported negative or positive tests. The model and details of calculation of the basic reproduction number $\Rnum$ are presented in the appendixs.

@@

In order to link to more applied models\david{what does that mean?  what is an ``applied model''?}, we constructed this model so we could specify the total \percap testing rate. We do this by defining the weighted size of the testing pool $W = W_S S_u + W_I I_u + W_R R_u$, and calculating a scaling parameter for testing as:
\begin{equation}
\label{sigma}
\sigma = \frac{\rho N_0}{W}.
\end{equation}
Thus, the \percap testing rate for compartment $X$ is $F_X=\sigma W_X$, where $X \in \{S,I,R\}$. 
For a high-sensitivity test, infected people typically flow through to the ``confirmed positive'' ($I_c$, $R_c$) compartments and are thus unavailable for further testing.  Over the course of the epidemic, a fixed testing rate as specified in \eqref{sigma} can (if large enough) exhaust the pool of people available for testing, leading to a singularity when no one is left untested.  Although this phenomenon does not affect our analysis of $\Rnum$, it can affect the temporal dynamics (we discuss an adjustment to the model that solves this problem in the appendix).

\david{A comment on notation:  You haven't actually defined $S_u$, $I_u$ etc.  In any case, I urge you to use macros and use Roman font for subscripts that are textual as opposed to variables.  So $S_{\rm u}$ and $I_{\rm p}$ etc.  Then it is clear the subscript isn't a variable with a value.}\ali{will do}

The classical SIR model is based on the following implicit assumptions; well-mixed population, homogeneity of the population\david{What's the difference btw ``well-mixed'' and ``homoegeneous''?}\ali{ non structured eg age, ok?}, exponentially distributed duration of infection and large population size (see, e.g., \cite{keeling2011modeling}). In addition to these standard assumptions, our model, \ref{model}, assumes: (i) there is a single force of infection, $\Lambda$, defined as follows
\begin{equation}
\label{Lambda}
\Lambda=\beta \frac{(I_u+\eta_w I_n+\eta_w I_p+ \eta_c I_c)}{N_0},
\end{equation}
across all susceptible pools (see Table~\ref{tab:params} for details\david{there are no details in the table; you need to do a better job of explaining what's going on}), (ii) $\eta_c \leq \eta_w$, i.e., the individuals awaiting test results have a higher transmission probability than the reported individuals. For this analysis, we also assume a perfectly specific test ($p_S=0$). This last assumption combined with the assumption that no individual is in waiting-for-\emph{positive} or \emph{confirmed positive} compartments, i.e., $S_p(0)=S_c(0)=0$, reduces the model to 10 equations (equations c and d  in \eqref{model} are eliminated).

@@

The Disease-Free Equilibrium (DFE) for the SIR model, Eqs. \eqref{model}, is given by solving the coupled system including $S_u+S_n=N_0$ and $F_S S_u-\omega S_n=0$.
\david{you mean by solving the algebraic equations obtain by setting the DEs to 0.  not clear why you bother to mention two of these algebraic eqns.}
The DFE is
\begin{equation}
\label{dfe}
S_n^*= \frac{\rho}{\omega} N_0, \ S_u^*= N_0-S_n^*, \text{~and~} I_j=R_j=0 \ \text{for all }j.
\end{equation}
\david{above you had j in Roman font even though it *is* a variable, so I changed to $j$}

The basic reproduction number, $\Rnum$, was calculated by using the next generation matrix method developed by \cite{van2002reproduction}. $\Rnum$ is
\begin{equation}
\label{R0}
\Rnum= \frac{\beta}{\gamma} (A \times S_u^* + B \times S_n^*) \times C, 
\end{equation}
where
\begin{equation}
\begin{aligned}
\label{eq:abc}
A=& \gamma(\omega+\gamma) + (\gamma \eta_w + \omega \eta_c p_I) F_I, \\
B=& \big(\omega+(F_I+\gamma)\eta_w\big) \gamma+\frac{(\eta_w \gamma+ \eta_c\omega) \omega p_I F_I }{\omega+\gamma}, \\ 
C=& \frac{1}{N_0 (\gamma(\omega+\gamma)+F_I(\gamma+\omega p_I))}.
\end{aligned}
\end{equation}
\david{put $\beta/\gamma$ in front of the expression in eq \eqref{R0}, so it is clear that the rest is a modification of the the simplest expression.  It would be worth spending some time trying to write the expression in a way that helps the reader understand where the different factors come from.}
Further details of derivation of $\Rnum$ are provided in the appendix section.

@@

The explicit formula for the basic reproduction number, $\Rnum$ \eqref{R0}, provides an opportunity to study the influence of changes in the underlying parameters on the critical index of epidemic dynamics. We are interested in understanding the effect of parameters that can be realistically controlled by changing in isolation, \percap testing intensity and test resulting, i.e., $\eta_c$ and $\eta_w$, $\rho$ and $\omega$, respectively. The following Proposition is the direct result of taking partial derivative of $\Rnum$ \eqref{R0} with respect to selected parameters. Note that we used Taylor approximation of $\Rnum$ at $\rho=0$ due to the complexity of the expression and thus inconclusive.\david{in that case it would be good to establish numerically that the relationships given in the proposition are valid more generally, at least for reasonable parameter ranges}
See the appendix for details.

\david{make sure to use $\Rnum$ not $\Rnum$ below (and often in the supp material section).  More importantly, you really need to state in words what this proposition is saying.  The reader has to check the param table and think it through themselves.}
\begin{proposition}
\label{prop1}
Using the expression of $\Rnum$ \eqref{R0},
\begin{enumerate}
\item \label{p1:eta}
$\partial{\Rnum}/\partial{\eta_c} \geq 0$ and $\partial{\Rnum}/\partial{\eta_w} \geq 0$. 
\item \label{p1:rho}
$\partial{\Rnum}/\partial{\rho} \leq 0$ when $\rho \approx 0$.
\item \label{p1:omega}
$\partial{\Rnum}/\partial{\omega}$ can be positive or negative when $\rho \approx 0$.
\end{enumerate}
\end{proposition}

@@

We incorporated the compartment-specific relative testing weights, $W_S$, $W_I$ and $W_R$, to model random testing and \emph{targeted} testing strategies. Here, in the case of targeted testing and for the simplicity and illustration purposes, we assumed that infected and recovered individuals are tested at three times the \percap rate of susceptible individuals, thus $W_I/W_S=3$ and $W_R=W_I$. Note that we have not specified a methodology to assign particular relative testing weights corresponding to a particular targeted testing scenario. 
This part needs to be developed further in future work. 
Modeling different targeted testing strategies, equivalently test-specific testing weights in our framework, requires prior information of the conditional probabilities of getting tested for people in a given compartment. 
This can be implied when we would like to quantify and compare the effect of different levels of test focus for infectious people on the basic reproduction number $\Rnum$, and conclude about the disease spread management. For example, when people are tested for "screening", the individuals with potential higher mobility, eg. people who are getting on flights, get more tested and thus the coresponding heavier testing weight is assigned than people awaiting for a surgery and are probably going to stay in a long-term care facility and consequently less mobile and more isolated to begin with. With our model, we would be able to compare the sensitivity of the disease dynamics, through $\Rnum$, with respect to testing processes.

{\bf The \percap testing intensity, $\rho$;}
Proposition \ref{prop1} part \ref{p1:rho} and also Figure \ref{pan} indicates that increasing testing intensity $\rho$ reduces $\Rnum$. This is sensible since as people are moved to test compartments, namely $X_p$ and $X_n$ for $X \in \{S,I,R\}$, they may partially or fully self-isolate. Also individuals who have tested positive, namely $X_c$ compartment for $X \in \{S,I,R\}$, are highly likely to self-isolate. The higher probability of being subject to isolation, the lower the force of infection \eqref{Lambda} and the lower $\Rnum$ will become. While our simulation supported this result, analytically it appeared to be hard to conclude this due to the complexity of $\Rnum$ expression.

{\bf The potential advantage of slow test reporting, or favorable-delay-reporting;}
Individuals are highly likely to fully self-isolate when they are either awaiting the test results or they are reported, thus reducing or eliminating their potentially infectious contacts. Thus, the faster the test reporting rate, $\omega$, the shorter these individuals stay in the ``\emph {safe}'' awaiting-confirmed compartments, namely $X_n$, $X_p$ and $X_c$ for $X \in \{S,I,R\}$, and the more they get involved in the infection process.
This advantage of slow test reporting is real, and neglected. 
We also compare to an individual-level advantage of fast tests: people who test positive may be even more careful.
In our model analysis, Proposition \ref{prop1} part \ref{p1:omega} is describing this potential advantage. 
It states that returning test results more rapidly, i.e., increasing the rate $\omega$, does not necessarily lower the reproduction number $\Rnum$; whether increasing $\omega$ lowers $\Rnum$ depends on the precise combination of model parameters  including test reporting rate, testing strategies represented by compartment-specific testing weights, test sensitivity and specificity, and the level of isolation. 
Specifically, in the case of perfect isolation, i.e., when $\eta_w=0$ and $\eta_c=0$, $\Rnum$ may increase as the test reporting process becomes faster. This can be seen from expression \eqref{Rom}.
Another example of this favorable delay in $\omega$ could be when the test being employed produces many false negatives. Because many infected asymptomatic individuals will believe they are uninfected/uninfectious, thus may unknowingly spread the virus to many others. Again the delay in the test reporting rate keeps these individuals in the ``\emph {safe}'' awaiting-confirmed compartments.

we are missing out on
community-level advantages of fast testing: better assessment of the
situation, identification of hot spots, contact-tracing, etc.
The \percap testing intensity of CIVID-19 after about a year from the first case reported in December 2019, is still low (($\rho \approx 0$) in our model). In near future new test kits may be widely accessible, our model provide insights in this case. 
In particular, if a cheap test can identify on average more infected individuals as an expensive test, then our model predicts that the cheap test will lower $\Rnum$ more. In contrast, if an expensive test can identify on average more infected individuals, it will not necessarily lower $\Rnum$ more than the cheap test. The use of tests cheaper than RT-PCR has been proposed as a potential strategy for containing the COVID pandemic. While cheaper tests may be less sensitive and reliable than RT-PCR, they allow for broader and more intense testing. Using our Taylor approximation of $\Rnum$ near $\rho = 0$, we examined what circumstances (i.e., model parameters) make the use of one test more favourable than another, and give a complete description through inequality \ref{eq:rho1vsrho2}. In general, we found that the expensive test tends to more effectively lower $\Rnum$ when (a) individuals who test positive self-isolate much more than individuals who are waiting for their test result, (b) the time it takes to return tests is much shorter than the mean infectious period, and (c) the testing intensity is much greater for infected individuals than susceptible individuals.

\ali{ not sure if we want these 2 following parageraphs, or shrink it into afew short sentences?}
In addition to the favorable-delay-reporting observation of our model \eqref{model}, the model enables us to quantify the amount of delay required in the test reporting process as a strategy to reduce $\Rnum$.
To give a biological interpretation, we describe the qualitative trends predicted by the inequality \eqref{eq:necsuf}. To summarise, returning test results more rapidly tends to be favourable (i.e., reduces $\Rnum$) when (i) \emph{Confirmed-positive} individuals, lower their contact much more than individuals who are waiting for their test results (i.e., $\eta_c \ll \eta_w$), (ii) the test is highly sensitive (i.e., $p_I$ is close to 1), and (iii) the targeted testing strategy is used (i.e., $W_i \gg W_s$).

@@

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage
% \widetext
\begin{center}
\textbf{\large Appendix}
\end{center}
%%%%%%%%%% Merge with supplemental materials %%%%%%%%%%
%%%%%%%%%% Prefix a "S" to all equations, figures, tables and reset the counter %%%%%%%%%%
\setcounter{equation}{0}
\setcounter{figure}{0}
\setcounter{table}{0}
\setcounter{page}{1}
\makeatletter
\renewcommand{\theequation}{A\arabic{equation}}
\renewcommand{\thefigure}{A\arabic{figure}}
\renewcommand{\bibnumfmt}[1]{[A#1]}
\renewcommand{\citenumfont}[1]{A#1}
%%%%%%%%%% Prefix a "S" to all equations, figures, tables and reset the counter %%%%%%%%%%

@@

The use of tests cheaper than RT-PCR has been proposed as a potential strategy for containing the \covid pandemic. While cheaper tests may be less sensitive and reliable than RT-PCR, they allow for broader and more intense testing. In the analysis below, we compare the $\Rnum$ predicted by our model depending on the testing strategy.
